---
title: "Defensive Modeling"
author: "Claire Dudley"
date: "2023-12-01"
output: html_document
---


```{r}
library(tidyverse)
library(nflfastR)
library(nflreadr)
library(dplyr)
library(tictoc)
library(zoo)
library(lme4)
library(lmerTest)
```

```{r}
pbp_2020 <- nflreadr::load_pbp(2020)
all <-
  load_player_stats(2020)
```

```{r}
all2 <- nflfastR::load_pbp(2020) %>%
  dplyr::filter(season_type == "REG") %>%
  nflfastR::calculate_player_stats_def(weekly = TRUE)
```

# Defensive Players

```{r}
all3 <- left_join(all2, all, by = c('season','player_id', 'player_name', 'player_display_name', 'position', 'position_group', 'week'))
```



```{r}
defense_group <-
  all3 %>%
  filter(position_group == 'DB' | position_group == 'LB')

defense_group <-
  defense_group %>%
  select(-c('headshot_url.x', 'headshot_url.y'))
  

defense_group <- 
  defense_group %>%
  mutate_if(is.numeric, list(~replace_na(., 0)))
```

  So I currently have the fantasy points that individual dbacks and linebackers earned for their respective teams.
  I am going to have to sum rows by team (?), to predict team's earned fantasy points by actions such as sacks.
  
Equation for Defensive Team Fantasy Points:

  1 * sack + 2 * blocked kick + 2 * INT + 2 * Fumble Recovered + 2 * Safety + (-5 to 5) * Points Allowed + (-7 to 5) * Yards Allowed
  
Let's see what we've got: 
* I don't have special teams right now and I am going to start by not dealing with that (Blocked Kicks)
* Starting with just defensive calculations
* I know touchdowns would be points but how many?
* Points Allowed and Yards Allowed will have to be handled

```{r}
defense_group <-
  defense_group %>%
  mutate(defense_total = def_sacks + 2 * interceptions + 
           2 * def_fumble_recovery_own + 2 * def_safety)

defense_group$defense_total
```


# Making selective dataset
```{r}
def_select <-
  defense_group %>%
  dplyr::select('season', 'week',
                'team', 'def_sacks', 'interceptions', 'def_fumble_recovery_own', 'def_safety', 'fantasy_points', 'defense_total')
```

# Adding specific players' touchdowns/points
hopefully this is accurate but not many defensive players have anything for the fantasy_points variable
```{r}
def_select <-
  def_select %>%
  mutate(fantasy_total = fantasy_points + defense_total)
```

# Need to sum rows to get team values each week
```{r}
def_team_stats <- def_select %>%
  group_by(season, week, team) %>%
  summarise(across(everything(), sum),
            .groups = 'drop') %>%
  as.data.frame()
```

# OK now points against. 
# Offensive team statistics will be how I do this

```{r}
offensive_team_stats <- 
  nflfastR::load_pbp(2020) %>%
  nflfastR::calculate_player_stats(weekly = TRUE) %>%
  filter(!position_group %in% c('DB', 'LB'),
         season_type == 'REG')
```

```{r}
offense_sel <-
  offensive_team_stats %>%
  select('recent_team', 'season', 'week', 'passing_yards','passing_tds', 'rushing_yards', 'rushing_tds', 
         'receiving_yards', 'receiving_tds')
```


# total points from touchdowns and total yards
```{r}
offense_sel <-
  offense_sel %>%
  mutate(td_points = rushing_tds * 7 + passing_tds * 7,
         total_yards = rushing_yards + passing_yards)
```

# summing by team

```{r}
off_team_stats <- offense_sel %>%
  group_by(season, week, recent_team) %>%
  summarise(across(everything(), sum),
            .groups = 'drop') %>%
  as.data.frame()
```


# preparing to join

```{r}
off_team_stats <-
  off_team_stats %>%
  rename(posteam = recent_team)

def_team_stats <-
  def_team_stats %>%
  rename(defteam = team)
```

pbp data
```{r}
pbp_2020_join <-
  pbp_2020 %>%
  dplyr::select(week, posteam, defteam)
```

```{r}
off_team_total <-
  left_join(off_team_stats, pbp_2020_join, by = c('posteam', 'week')) %>% distinct()
```

```{r}
def_team_stats_nec <- 
  def_team_stats %>%
  dplyr::select('week', 'defteam', 'fantasy_total')
```

```{r}
combo <-
  left_join(off_team_total, def_team_stats_nec, by = c('week', 'defteam'))
```

# Kicking/field goal points

```{r}
kicking_stats <- 
  nflfastR::load_pbp(2020) %>%
  nflfastR::calculate_player_stats_kicking(weekly = TRUE) 
```

# Calculating points from field goals

```{r}
kicking_stats <- 
  kicking_stats %>%
  mutate(fg_points = fg_made * 3) %>%
  dplyr::select('week', 'team', 'fg_points')

kicking_stats <-
  kicking_stats %>%
  rename(posteam = team)

kicking_stats[is.na(kicking_stats)] <- 0
```

# Joining kicking info

```{r}
final_set <- left_join(combo, kicking_stats, by = c('week', 'posteam'))
```

```{r}
final_set <-
  final_set %>%
  mutate(off_points = fg_points + td_points)
```

# defensive points for how many points were given up

```{r}
final_set <-
  final_set %>%
  mutate(points_allowed = case_when(off_points < 1 ~ 10,
                                    off_points > 0 & off_points < 7 ~ 7,
                                    off_points > 6 & off_points < 14 ~ 4,
                                    off_points > 13 & off_points < 21 ~ 1,
                                    off_points > 20 & off_points < 28 ~ 0,
                                    off_points > 27 & off_points < 35 ~ -1, 
                                    off_points > 34 ~ -4))
```

```{r}
final_set <-
  final_set %>%
  mutate(points = points_allowed + fantasy_total)
```

# 5 week rolling averages

```{r}
final_set <-
  final_set %>%
  arrange(defteam, week) %>%
  group_by(defteam) %>%
  mutate(avg_yards = lag(rollapplyr(total_yards, 5, mean, partial = TRUE)),
         avg_tdpoints = lag(rollapplyr(td_points, 5, mean, partial = TRUE)),
         avg_offpoints = lag(rollapplyr(off_points, 5, mean, partial = TRUE)),
         avg_points = lag(rollapplyr(points, 5, mean, partial = TRUE))
         )

final_set <-
  final_set %>%
  filter(!is.na(avg_yards))
```

# OK Modeling now

linear
```{r}
linear_mod <- lm(points ~ avg_yards + avg_points + avg_offpoints, data = final_set)
summary(linear_mod)

predictions <- predict(linear_mod, newdata = final_set, allow.new.levels = TRUE)
mse <- mean((predictions - final_set$points) ^ 2, na.rm = TRUE)
mse
rmse <- sqrt(mse)
rmse
```

mixed effects

```{r}
def_lmer <- lmer(points ~ avg_yards + avg_offpoints + (1|posteam),
                data = final_set)

summary(def_lmer)
```

testing
```{r}
predictions <- predict(def_lmer, newdata = final_set, allow.new.levels=TRUE)

mse <- mean((predictions - final_set$points) ^ 2, na.rm = TRUE)
mse
rmse <- sqrt(mse)
rmse
```

# Individual Defense

```{r}
defense_group <-
  defense_group %>%
  mutate(individual_fantasy = def_tackles_solo + 0.5 * def_tackle_assists + def_tackles_for_loss + def_qb_hits + 2 * def_sacks + 3 * interceptions + def_pass_defended + 3 * def_fumbles_forced + 6 * def_tds)
```

```{r}
def_sel2 <-
  defense_group %>%
  dplyr::select('season', 'week',
                'team', 'def_sacks', 'interceptions', 'def_fumbles_forced', 'def_tackles_solo', 'def_tackles_for_loss',
                'def_qb_hits','individual_fantasy')

def_sel2  <-
  def_sel2 %>%
  rename(defteam = team)
```

```{r}
indiv_defense <-
  left_join(def_sel2, off_team_total, by = c("defteam", "week"))
```

# 5 week rolling averages

```{r}
indiv_defense <-
  indiv_defense %>%
  arrange(defteam, week) %>%
  group_by(defteam) %>%
  mutate(avg_sacks = lag(rollapplyr(def_sacks, 5, mean, partial = TRUE)),
         avg_ints = lag(rollapplyr(interceptions, 5, mean, partial = TRUE)),
         avg_tackles = lag(rollapplyr(def_tackles_solo, 5, mean, partial = TRUE)),
         avg_qbhits = lag(rollapplyr(def_qb_hits, 5, mean, partial = TRUE))
         )

indiv_defense <-
  indiv_defense %>%
  filter(!is.na(avg_sacks))
```


# Test Data
first team defense:

```{r}
all_23 <-
  load_player_stats(2023)
all2_23 <- nflfastR::load_pbp(2023) %>%
  dplyr::filter(season_type == "REG") %>%
  nflfastR::calculate_player_stats_def(weekly = TRUE)
```

```{r}
all3_23 <- left_join(all2_23, all_23, by = c('season','player_id', 'player_name', 'player_display_name', 'position', 'position_group', 'week'))
```

```{r}
defense_group_23 <-
  all3_23 %>%
  filter(position_group == 'DB' | position_group == 'LB')

defense_group_23 <-
  defense_group_23 %>%
  select(-c('headshot_url.x', 'headshot_url.y'))
  

defense_group_23 <- 
  defense_group_23 %>%
  mutate_if(is.numeric, list(~replace_na(., 0)))
```

```{r}
defense_group_23 <-
  defense_group_23 %>%
  mutate(defense_total = def_sacks + 2 * interceptions + 
           2 * def_fumble_recovery_own + 2 * def_safety)

defense_group_23$defense_total
```

```{r}
def_select_23 <-
  defense_group_23 %>%
  dplyr::select('season', 'week',
                'team', 'def_sacks', 'interceptions', 'def_fumble_recovery_own', 'def_safety', 'fantasy_points', 'defense_total')
```

```{r}
def_select_23 <-
  def_select_23 %>%
  mutate(fantasy_total = fantasy_points + defense_total)
```

```{r}
def_team_stats_23 <- def_select_23 %>%
  group_by(season, week, team) %>%
  summarise(across(everything(), sum),
            .groups = 'drop') %>%
  as.data.frame()
```

# OK now points against. 
# Offensive team statistics will be how I do this

```{r}
offensive_team_stats_23 <- 
  nflfastR::load_pbp(2023) %>%
  nflfastR::calculate_player_stats(weekly = TRUE) %>%
  filter(!position_group %in% c('DB', 'LB'),
         season_type == 'REG')
```

```{r}
offense_sel_23 <-
  offensive_team_stats_23 %>%
  select('recent_team', 'season', 'week', 'passing_yards','passing_tds', 'rushing_yards', 'rushing_tds', 
         'receiving_yards', 'receiving_tds')
```


# total points from touchdowns and total yards
```{r}
offense_sel_23 <-
  offense_sel_23 %>%
  mutate(td_points = rushing_tds * 7 + passing_tds * 7,
         total_yards = rushing_yards + passing_yards)
```

# summing by team

```{r}
off_team_stats_23 <- offense_sel_23 %>%
  group_by(season, week, recent_team) %>%
  summarise(across(everything(), sum),
            .groups = 'drop') %>%
  as.data.frame()
```


# preparing to join

```{r}
off_team_stats_23 <-
  off_team_stats_23 %>%
  rename(posteam = recent_team)

def_team_stats_23 <-
  def_team_stats_23 %>%
  rename(defteam = team)
```

pbp data
```{r}
pbp_2023_join <-
  pbp_2023 %>%
  dplyr::select(week, posteam, defteam)
```

```{r}
off_team_total_23 <-
  left_join(off_team_stats_23, pbp_2023_join, by = c('posteam', 'week')) %>% distinct()
```

```{r}
def_team_stats_nec_23 <- 
  def_team_stats_23 %>%
  dplyr::select('week', 'defteam', 'fantasy_total')
```

```{r}
combo_23 <-
  left_join(off_team_total_23, def_team_stats_nec_23, by = c('week', 'defteam'))
```

# Kicking/field goal points

```{r}
kicking_stats_23 <- 
  nflfastR::load_pbp(2023) %>%
  nflfastR::calculate_player_stats_kicking(weekly = TRUE) 
```

# Calculating points from field goals

```{r}
kicking_stats_23 <- 
  kicking_stats_23 %>%
  mutate(fg_points = fg_made * 3) %>%
  dplyr::select('week', 'team', 'fg_points')

kicking_stats_23 <-
  kicking_stats_23 %>%
  rename(posteam = team)

kicking_stats_23[is.na(kicking_stats_23)] <- 0
```

# Joining kicking info

```{r}
final_set_23 <- left_join(combo_23, kicking_stats_23, by = c('week', 'posteam'))
```

```{r}
final_set_23 <-
  final_set_23 %>%
  mutate(off_points = fg_points + td_points)
```

# defensive points for how many points were given up

```{r}
final_set_23 <-
  final_set_23 %>%
  mutate(points_allowed = case_when(off_points < 1 ~ 10,
                                    off_points > 0 & off_points < 7 ~ 7,
                                    off_points > 6 & off_points < 14 ~ 4,
                                    off_points > 13 & off_points < 21 ~ 1,
                                    off_points > 20 & off_points < 28 ~ 0,
                                    off_points > 27 & off_points < 35 ~ -1, 
                                    off_points > 34 ~ -4))
```

```{r}
final_set_23 <-
  final_set_23 %>%
  mutate(points = points_allowed + fantasy_total)
```

# 5 week rolling averages

```{r}
final_set_23 <-
  final_set_23 %>%
  arrange(defteam, week) %>%
  group_by(defteam) %>%
  mutate(avg_yards = lag(rollapplyr(total_yards, 5, mean, partial = TRUE)),
         avg_tdpoints = lag(rollapplyr(td_points, 5, mean, partial = TRUE)),
         avg_offpoints = lag(rollapplyr(off_points, 5, mean, partial = TRUE)),
         avg_points = lag(rollapplyr(points, 5, mean, partial = TRUE))
         )

final_set_23 <-
  final_set_23 %>%
  filter(!is.na(avg_yards))
```

# OK Modeling now

```{r}
def_lmer <- lmer(points ~ avg_yards + avg_offpoints + (1|posteam),
                data = final_set)
```

testing
```{r}
predictions <- predict(def_lmer, newdata = final_set_23, allow.new.levels=TRUE)

mse <- mean((predictions - final_set_23$points) ^ 2, na.rm = TRUE)
mse
rmse <- sqrt(mse)
rmse
```

# Individual Players

```{r}
defense_group_23 <-
  defense_group_23 %>%
  mutate(individual_fantasy = def_tackles_solo + 0.5 * def_tackle_assists + def_tackles_for_loss + def_qb_hits + 2 * def_sacks + 3 * interceptions + def_pass_defended + 3 * def_fumbles_forced + 6 * def_tds)
```

```{r}
def_sel2_23 <-
  defense_group_23 %>%
  dplyr::select('season', 'week',
                'team', 'def_sacks', 'interceptions', 'def_fumbles_forced', 'def_tackles_solo', 'def_tackles_for_loss',
                'def_qb_hits','individual_fantasy')

def_sel2_23  <-
  def_sel2_23 %>%
  rename(defteam = team)
```

```{r}
indiv_defense_23 <-
  left_join(def_sel2, off_team_total, by = c("defteam", "week"))
```

# 5 week rolling averages

```{r}
indiv_defense_23 <-
  indiv_defense_23 %>%
  arrange(defteam, week) %>%
  group_by(defteam) %>%
  mutate(avg_sacks = lag(rollapplyr(def_sacks, 5, mean, partial = TRUE)),
         avg_ints = lag(rollapplyr(interceptions, 5, mean, partial = TRUE)),
         avg_tackles = lag(rollapplyr(def_tackles_solo, 5, mean, partial = TRUE)),
         avg_qbhits = lag(rollapplyr(def_qb_hits, 5, mean, partial = TRUE))
         )

indiv_defense_23 <-
  indiv_defense_23 %>%
  filter(!is.na(avg_sacks))
```

modeling

```{r}
indiv_def_lmer <- lmer(individual_fantasy ~ avg_sacks + avg_ints + avg_tackles + avg_qbhits + (1|posteam),
                       data = indiv_defense)
summary(indiv_def_lmer)
```


testing
```{r}
predictions <- predict(indiv_def_lmer, newdata = indiv_defense_23, allow.new.levels=TRUE)

mse <- mean((predictions - indiv_defense_23$individual_fantasy) ^ 2, na.rm = TRUE)
mse
rmse <- sqrt(mse)
rmse
```




